### 2023-04-04 chapter1/chapter1.tex
Fixed bad path in to new realignment figure
Added clarification to Gaussian process sequence weighting
Fixed calling missing data phylo-HMM as deletion phylo-HMM

\begin{figure}[h!]
\includegraphics[width=\textwidth]{realignment/out/merged.png}
\centering
\caption{\textbf{Alignment with long poorly supported segments.}
The alignments of representative sequences in orthologous groups 0167 \textbf{(A)}, 2770 \textbf{(B)}, and 23D9 \textbf{(C)} before and after refinement.}
\label{fig:realignment}
\end{figure}

@@

All HMM algorithms were implemented with custom code which is available as the package Homomorph on PyPI. The insertion phylo-HMM was trained on 47,387 manually labeled columns in 14 alignments, and the deletion phylo-HMM was trained on 67,001 manually labeled positions in 23 sequences in 11 unique alignments (Fig.~\ref{sfig:insertion_training}, Fig.~\ref{sfig:missing_training}). Because maximum-likelihood estimation of the model parameters yielded posterior decoding curves which toggled between hidden states too rapidly, the models were instead trained discriminatively~\cite{Krogh1999}. The difference, briefly, is maximum-likelihood estimation finds the parameters that best reproduce the observed distributions whereas discriminative training finds the parameters that minimize prediction error. Discriminatively trained models typically perform better in practice since real-world data are rarely fully described by the distribution specified by the model.

@@

 States 2 and 3 have distinct characteristics which required different trimming strategies. Regions with a high probability of state 3 were handled first. Because the long poorly supported segments in state 3 regions were sometimes aligned to highly conserved columns or short segments in other sequences, trimming columns entirely would remove these segments from the other sequences even if they would not qualify as long and poorly supported themselves. Thus, regions with high state 3 probabilities were trimmed at the level of individual sequences rather than entire columns using the following method. First, the probability of state 3 for columns with a gap profile value less than or equal to 0.1 was set to 0 to break long poorly supported segments aligned to highly conserved columns into separate regions. Regions were defined with the previously described algorithm using high and low cutoffs of 0.75 and 0.01, respectively. The outer and inner derivative cutoffs were both 0.001. The mean number of non-gap symbols in a region was calculated using Gaussian process sequence weighting. (The sequences with the five most non-gap symbols were also excluded to not bias the estimate with long poorly supported segments.) The final mean $\mu$ was taken as the minimum of this value and two. A cutoff \textit{k}, derived from a geometric model of the number of non-gap symbols and a significance level $\alpha$, was calculated using the equation $k = \log(\alpha) / \log(1 - p) - 1$ where $p = \frac{1}{\mu + 1}$ and $\alpha = 0.01$. Any sequence whose number of non-gap symbols in the region equaled or exceeded this value was trimmed by replacing all non-gap symbols with gaps.

>>>>

\begin{figure}[h!]
\includegraphics[width=\textwidth]{realignment/out/realignment.png}
\centering
\caption{\textbf{Alignment with long poorly supported segments.}
The alignments of representative sequences in orthologous groups 0167 \textbf{(A)}, 2770 \textbf{(B)}, and 23D9 \textbf{(C)} before and after refinement.}
\label{fig:realignment}
\end{figure}

@@

All HMM algorithms were implemented with custom code which is available as the package Homomorph on PyPI. The insertion phylo-HMM was trained on 47,387 manually labeled columns in 14 alignments, and the missing data phylo-HMM was trained on 67,001 manually labeled positions in 23 sequences in 11 unique alignments (Fig.~\ref{sfig:insertion_training}, Fig.~\ref{sfig:missing_training}). Because maximum-likelihood estimation of the model parameters yielded posterior decoding curves which toggled between hidden states too rapidly, the models were instead trained discriminatively~\cite{Krogh1999}. The difference, briefly, is maximum-likelihood estimation finds the parameters that best reproduce the observed distributions whereas discriminative training finds the parameters that minimize prediction error. Discriminatively trained models typically perform better in practice since real-world data are rarely fully described by the distribution specified by the model.

@@

 States 2 and 3 have distinct characteristics which required different trimming strategies. Regions with a high probability of state 3 were handled first. Because the long poorly supported segments in state 3 regions were sometimes aligned to highly conserved columns or short segments in other sequences, trimming columns entirely would remove these segments from the other sequences even if they would not qualify as long and poorly supported themselves. Thus, regions with high state 3 probabilities were trimmed at the level of individual sequences rather than entire columns using the following method. First, the probability of state 3 for columns with a gap profile value less than or equal to 0.1 was set to 0 to break long poorly supported segments aligned to highly conserved columns into separate regions. Regions were defined with the previously described algorithm using high and low cutoffs of 0.75 and 0.01, respectively. The outer and inner derivative cutoffs were both 0.001. The mean number of non-gap symbols in a region was calculated using Gaussian process sequence weighting as described in the section on selecting representative sequences. (The sequences with the five most non-gap symbols were also excluded to not bias the estimate with long poorly supported segments.) The final mean $\mu$ was taken as the minimum of this value and two. A cutoff \textit{k}, derived from a geometric model of the number of non-gap symbols and a significance level $\alpha$, was calculated using the equation $k = \log(\alpha) / \log(1 - p) - 1$ where $p = \frac{1}{\mu + 1}$ and $\alpha = 0.01$. Any sequence whose number of non-gap symbols in the region equaled or exceeded this value was trimmed by replacing all non-gap symbols with gaps.
